# Analysis of Storm Affects on Population Health and Economy in USA

## Synopsis
In this report we aim to describe the affects of storms in USA on population health and economy by analyzing storm data from 1950 to 2011. Using the storm data obtained from  U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database, we present type of storm related events that are responsible for most negative affects. Top 25 such storm events are presented. Additionally individual events are grouped into event classes and affects of those are also presented.   


## Data Processing
### Analysis
We postulate that sum aggregates of damages will provide a meaninful measure for our analysis.  

Population health (HEALTHDAMAGES) for each event is obtained after cumulative sum of reported injuries and fatilities for each year between 1950 - 2011. Economic affect (DMGCASH) for each event is obtained after cumlative sum of crop and property damage and after applying appropiate multipler. The results are then sorted and top 25 events and their values are presented.  

Event Class affects is done similary except each storm data event is first grouped into a class and then the sums are obtained for each class. Please see MakeGroups() function (below) 

### Loading
The data used for this analysis is available at this URL https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2.  

```{r, echo=TRUE, cache=TRUE}
#repdata-data-StormData.csv.bz2 is obtained from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
storm.data <- read.csv("repdata-data-StormData.csv.bz2")
```

### Processing Health Affects
Subset the storm data to extract health affects
```{r, echo=TRUE,fig.height=10, fig.width=10}
library(ggplot2)
library(data.table)
library(xtable)
sd.health <- storm.data[ storm.data$FATALITIES > 0 | storm.data$INJURIES > 0, 
                         c("EVTYPE", "INJURIES", "FATALITIES")]
sd.health$EVTYPE2 <- as.factor(toupper(sd.health$EVTYPE))    
```

Calculate health affects (HEALTHDAMAGES)
```{r, echo=TRUE,fig.height=10, fig.width=10,results='asis'}
sd.health$HEALTHDAMAGES <- sd.health$FATALITIES + sd.health$INJURIES;
sd.health.sum <- tapply(sd.health$HEALTHDAMAGES, sd.health$EVTYPE2, sum)
sd.health.sum <- data.frame(EVTYPE2=names(sd.health.sum), 
                            HEALTHDAMAGES=sd.health.sum, row.names=NULL)
sd.health.sum <- sd.health.sum[order(sd.health.sum$HEALTHDAMAGES, 
                                     decreasing=TRUE), ]
xt <- xtable(sd.health.sum)
print(xt, type="html", include.rownames=FALSE)

```

### Processing Economic Affects
Subset the storm data to extract economic affects
```{r,echo=TRUE, fig.height=10, fig.width=10}
sd.econ <- storm.data[storm.data$PROPDMG > 0 | storm.data$CROPDMG > 0, 
                         c("EVTYPE", "PROPDMG","PROPDMGEXP", "CROPDMG","CROPDMGEXP")]
sd.econ$EVTYPE2 <- as.factor(toupper(sd.econ$EVTYPE))    
```

Calculate economic damages (DMGCASH).  

The numeric values and its multiple are present in different columns. The multipler can be either character (e.g "H", "K", "M" or "B" for hunderds, thousands, millions or billions) or exponent values. We assume the multiplier == 1 when it does not fall into either of the above two classes. 
```{r,echo=TRUE,warning=FALSE, fig.height=10, fig.width=10,results='asis'}
sd.econ$PROPDMGEXP <- as.factor(toupper(as.character(sd.econ$PROPDMGEXP)))
sd.econ$CROPDMGEXP <- as.factor(toupper(as.character(sd.econ$CROPDMGEXP)))

dvals <- unique(union(sd.econ$PROPDMGEXP, sd.econ$CROPDMGEXP))
DMGEXP <- as.integer(dvals)
names(DMGEXP) <- dvals
DMGEXP[names(DMGEXP) %in% c("", "+", "-", "?")] <- 0
DMGEXP["H"] <- 2
DMGEXP["K"] <- 3
DMGEXP["M"] <- 6
DMGEXP["B"] <- 9

sd.econ$PROPDMGCASH <- sd.econ$PROPDMG * (10 ^ DMGEXP[sd.econ$PROPDMGEXP])
sd.econ$CROPDMGCASH <- sd.econ$CROPDMG * (10 ^ DMGEXP[sd.econ$CROPDMGEXP])
sd.econ$DMGCASH <- sd.econ$CROPDMGCASH + sd.econ$PROPDMGCASH
#sum(is.na(sd.econ$DMGCASH))
sd.econ.sum <- tapply(sd.econ$DMGCASH, sd.econ$EVTYPE2, sum)
sd.econ.sum <- data.frame(EVTYPE2=names(sd.econ.sum),
                          DMGCASH=sd.econ.sum, row.names=NULL)
sd.econ.sum <- sd.econ.sum[order(sd.econ.sum$DMGCASH, decreasing=TRUE), ]
xt <- xtable(sd.econ.sum)
print(xt, type="html", include.rownames=FALSE)
```

### Processing Event Class Affects

Events Classes are grouped as follows. Note that event classes can overlap.
```{r, echo=TRUE}
MakeGroups <- function (sd) {
    
    tornado         <- grep("TORNADO", sd$EVTYPE2, ignore.case=TRUE)
    heat            <- grep("HEAT|HYPERTHERMIA", sd$EVTYPE2, ignore.case=TRUE)
    snow.cold       <- grep("SNOW|COLD|HYPOTHERMIA|WINTER|HAIL|BLIZZARD|ICE|ICY|FREEZ|FROST|GLAZE|WINTRY", 
                            sd$EVTYPE2, ignore.case=TRUE)
    drought         <- grep("DROUGHT", sd$EVTYPE2, ignore.case=TRUE)
    flood           <- grep("FLOOD|TSUNAMI|WATER|FLD", sd$EVTYPE2, ignore.case=TRUE)
    lightning       <- grep("LIGHTNING", sd$EVTYPE2, ignore.case=TRUE)
    fire            <- grep("FIRE", sd$EVTYPE2, ignore.case=TRUE)
    rain            <- grep("RAIN|LANDSLIDE|MUDSLIDE|AVALANCH?E", sd$EVTYPE2, ignore.case=TRUE)
    fog             <- grep("FOG", sd$EVTYPE2, ignore.case=TRUE)
    coastal         <- grep("CURRENT|SURF|SEAS|WAVE|MARINE", sd$EVTYPE2, ignore.case=TRUE)
    
    #different storm categorized
    thunderstorm    <- grep("THUNDERSTORM|TSTM", sd$EVTYPE2, ignore.case=TRUE)
    wind            <- setdiff(grep("WIND", sd$EVTYPE2, ignore.case=TRUE), thunderstorm)
    hurricane       <- setdiff(grep("HURRICANE|TYPHOON", sd$EVTYPE2, 
                                    ignore.case=TRUE), thunderstorm)    
    blizzard        <- grep("BLIZZARD", sd$EVTYPE2, ignore.case=TRUE)
    storm           <- setdiff(grep("STORM", sd$EVTYPE2, ignore.case=TRUE),
                               union(thunderstorm, c(blizzard, hurricane)))
    accounted <- unique(c(tornado, heat, snow.cold, drought, flood, lightning, fire, rain, fog, coastal,
                          thunderstorm, wind, hurricane, blizzard, storm))
    others <- setdiff(1:nrow(sd), accounted)
    
    groups <- data.frame()
    groups <- rbind(groups, data.frame(group="TORNADO", EVTYPE2=unique(sd[tornado, c("EVTYPE2")])))
    groups <- rbind(groups, data.frame(group="HEAT", EVTYPE2=unique(sd[heat, c("EVTYPE2")])))
    groups <- rbind(groups, data.frame(group="SNOW/COLD", EVTYPE2=unique(sd[snow.cold, c("EVTYPE2")])))
    groups <- rbind(groups, data.frame(group="DROUGHT", EVTYPE2=unique(sd[drought, c("EVTYPE2")])))
    groups <- rbind(groups, data.frame(group="FLOOD", EVTYPE2=unique(sd[flood, c("EVTYPE2")])))
    groups <- rbind(groups, data.frame(group="LIGHTNING", EVTYPE2=unique(sd[lightning, c("EVTYPE2")])))
    groups <- rbind(groups, data.frame(group="FIRE", EVTYPE2=unique(sd[fire, c("EVTYPE2")])))
    groups <- rbind(groups, data.frame(group="RAIN", EVTYPE2=unique(sd[rain, c("EVTYPE2")])))
    groups <- rbind(groups, data.frame(group="FOG", EVTYPE2=unique(sd[fog, c("EVTYPE2")])))
    groups <- rbind(groups, data.frame(group="COASTAL", EVTYPE2=unique(sd[coastal, c("EVTYPE2")])))
    groups <- rbind(groups, data.frame(group="THUNDERSTORM", 
                                       EVTYPE2=unique(sd[thunderstorm, c("EVTYPE2")])))
    groups <- rbind(groups, data.frame(group="WIND", EVTYPE2=unique(sd[wind, c("EVTYPE2")])))
    groups <- rbind(groups, data.frame(group="HURRICANE", EVTYPE2=unique(sd[hurricane, c("EVTYPE2")])))
    groups <- rbind(groups, data.frame(group="BLIZZARD", EVTYPE2=unique(sd[blizzard, c("EVTYPE2")])))
    groups <- rbind(groups, data.frame(group="OTHER STORMS", EVTYPE2=unique(sd[storm, c("EVTYPE2")])))
    groups <- rbind(groups, data.frame(group="OTHER", EVTYPE2=unique(sd[others, c("EVTYPE2")])))
    groups
}   
```

Calculate health affects by storm event classes
```{r, echo=TRUE, fig.height=10, fig.width=10,results='asis'}
library(data.table)

sd.health.grp <- MakeGroups(sd.health)
sd.health.mrg <- merge(sd.health.sum, sd.health.grp, 
                       by.x="EVTYPE2", by.y="EVTYPE2", 
                       all.x=TRUE, all.y=FALSE)

sd.health.mrg.sum <- tapply(sd.health.mrg$HEALTHDAMAGES, sd.health.mrg$group, sum)
sd.health.mrg.sum <- data.frame(GROUP=names(sd.health.mrg.sum), 
                            HEALTHDAMAGES=sd.health.mrg.sum, row.names=NULL)
xt <- xtable(sd.health.mrg.sum)
print(xt, type="html", include.rownames=FALSE)
```

Calculate economic affects by storm event classes

```{r, echo=TRUE, fig.height=10, fig.width=10,results='asis'}
sd.econ.grp <- MakeGroups(sd.econ)
sd.econ.mrg <- merge(sd.econ.sum, sd.econ.grp, 
                       by.x="EVTYPE2", by.y="EVTYPE2", 
                       all.x=TRUE, all.y=FALSE)

sd.econ.mrg.sum <- tapply(sd.econ.mrg$DMGCASH, sd.econ.mrg$group, sum)
sd.econ.mrg.sum <- data.frame(GROUP=names(sd.econ.mrg.sum),
                              DMGCASH=sd.econ.mrg.sum, row.names=NULL)
xt <- xtable(sd.econ.mrg.sum)
print(xt, type="html", include.rownames=FALSE)

```

## Results

### Health Affects

Show top health affects
```{r, echo=TRUE,fig.height=10, fig.width=10}
sd.health.sum.top <- head(sd.health.sum, n=25)
sd.health.sum.top$HEALTHDAMAGES <- sd.health.sum.top$HEALTHDAMAGES / 1000

sdh.gp <- ggplot(sd.health.sum.top, aes(x=EVTYPE2, y=HEALTHDAMAGES)) +
    xlab("Events") + 
    ylab("Health affects in Thousands") + 
    ggtitle(paste("Top", nrow(sd.health.sum.top), 
                  "strom related health affects in USA from 1950 - 2011")) +
    geom_bar(stat="identity", fill=rainbow(n=length(sd.health.sum.top$EVTYPE2))) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5))

sdh.gp
```

### Economic Affects

Show top economic affects
```{r,echo=TRUE, fig.height=10, fig.width=10}
sd.econ.sum.top <- head(sd.econ.sum, n=25)
sd.econ.sum.top$DMGCASH <- sd.econ.sum.top$DMGCASH / (10^9)

sde.gp <- ggplot(sd.econ.sum.top, aes(x=EVTYPE2, y=DMGCASH)) +
    xlab("Events") + 
    ylab("Economic affects in Billions") + 
    ggtitle(paste("Top", nrow(sd.econ.sum.top), 
                  "strom related economic affects in USA from 1950 - 2011")) +
    geom_bar(stat="identity", fill=rainbow(n=length(sd.econ.sum.top$EVTYPE2))) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5))

sde.gp
```

### Event Class Affects

We use multiplot function, which is obtained from http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
```{r, echo=FALSE}
# Multiple plot function
# Obtained from : http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  require(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

Show health and economic affects by storm event classes 

```{r, echo=TRUE, fig.height=10, fig.width=10}
sd.health.mrg.sum$HEALTHDAMAGES <- sd.health.mrg.sum$HEALTHDAMAGES / 1000
sd.econ.mrg.sum$DMGCASH <- sd.econ.mrg.sum$DMGCASH / (10^9)

sdh.grp.gp <- ggplot(sd.health.mrg.sum, aes(x=GROUP, y=HEALTHDAMAGES)) +
    xlab("Event Classes") + 
    ylab("Health affects in Thousands") + 
    ggtitle("Health affects") +
    geom_bar(stat="identity", fill=rainbow(n=length(sd.health.mrg.sum$GROUP))) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5))

sde.grp.gp <- ggplot(sd.econ.mrg.sum, aes(x=GROUP, y=DMGCASH)) +
    xlab("Event Classes") + 
    ylab("Economic affects in Billions") + 
    ggtitle("Economic affects") +
    geom_bar(stat="identity", fill=rainbow(n=length(sd.econ.mrg.sum$GROUP))) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5))

multiplot(sdh.grp.gp, sde.grp.gp, cols=2)
```


## Conclusion

We have presented the types of storm related events that are responsible for most negative affects in USA from 1950 t0 2011 
